---
title: "Selecting 1968 Airphotos for the Neexdzii Kwah Floodplain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting 1968 Airphotos for the Neexdzii Kwah Floodplain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette walks through a real photo selection for the Neexdzii Kwah
(Upper Bulkley River) watershed in northwest BC. The goal: find every
available 1968 airphoto covering the modelled coho floodplain, so we can
understand what the riparian zone looked like before decades of land use
change. Was it rangeland? Old growth cottonwood forest? Wetland complex?
The airphotos are the ground truth.

## The two-tier AOI

diggs uses a two-tier area of interest:

1. **Watershed window** — configured in `data-raw/cache_data.R` before the app
   starts. This defines the regional extent: which photo centroids get downloaded
   from the BC Data Catalogue and cached locally. For Neexdzii Kwah, that's
   9,388 centroids spanning 1963--2019.

2. **Floodplain refinement** — a custom AOI uploaded or drawn inside the app.
   This narrows the selection to just the photos that cover your actual study
   area. Here, that's the 123.5 km^2^ coho floodplain delineated by
   [flooded](https://github.com/NewGraphEnvironment/flooded).

The watershed window is broad (you only run it once). The floodplain refinement
is precise (you iterate on it as your analysis evolves).

## Step 1: Delineate the floodplain

The floodplain AOI comes from `flooded::fl_valley_confine()` — a valley
confinement algorithm that uses a DEM, slope raster, stream network, and
precipitation to model where water can spread laterally.

The script `data-raw/vignette_neexdzii.R` runs this pipeline:

```{r floodplain}
library(flooded)

# Query coho streams (order >= 4) from bcfishpass
conn <- DBI::dbConnect(RPostgres::Postgres(),
  host = "localhost", port = 63333,
  dbname = "bcfishpass", user = "newgraph"
)
streams <- sf::st_read(conn, query = sql) |> sf::st_zm(drop = TRUE)
# 1,165 segments across 20 named streams

# Rasterize precipitation (mean annual precip from upstream area)
precip_r <- fl_stream_rasterize(streams, dem, field = "map_upstream")

# Run VCA
valleys <- fl_valley_confine(
  dem, streams,
  field = "upstream_area_ha",
  slope = slope,
  slope_threshold = 9,
  max_width = 2000,
  cost_threshold = 2500,
  flood_factor = 6,
  precip = precip_r,
  size_threshold = 5000,
  hole_threshold = 2500
)

# Polygonize and save as GeoJSON for diggs
valleys_poly <- fl_valley_poly(valleys)
sf::st_write(sf::st_transform(valleys_poly, 4326),
  "data/floodplain_neexdzii_co_4th_order.geojson"
)
```

This produces a single polygon covering 123.5 km^2^ of modelled floodplain
along 4th+ order coho streams.

## Step 2: Cache the watershed data

Before launching diggs, run `data-raw/cache_data.R` to download reference
layers and photo centroids from the BC Data Catalogue. The script is
parameterized by `blk` (blue line key) and `drm` (downstream route measure):

```{r cache}
# In data-raw/cache_data.R:
blk <- 360873822    # Bulkley River
drm <- 166030.4     # Neexdzii Kwa / Wedzin Kwa confluence

source("data-raw/cache_data.R")
# Caches 9,388 photo centroids (1963-2019), streams, railways, NTS grid
```

## Step 3: Launch diggs and select photos

```{r launch}
diggs::run_app()
```

In the app:

1. Switch AOI mode to **Custom (draw or upload)**
2. Upload `data/floodplain_neexdzii_co_4th_order.geojson`
3. Set year range to **1968--1968**
4. Click **Select** to run resolution-prioritized selection
5. Click **Download CSV** to export

## Step 4: What we got

The selection returned **40 photos** at two scales:

| Scale | Photos | Resolution |
|-------|--------|------------|
| 1:12,000 | 32 | High — individual trees visible |
| 1:31,680 | 8 | Medium — land cover classes distinguishable |

The 1:12,000 photos are the priority — they cover most of the floodplain at
a resolution where you can identify cottonwood stands, wetland boundaries,
and channel morphology. The 1:31,680 photos fill gaps where finer-scale
coverage doesn't exist.

```{r results}
sel <- read.csv("data/photo_selection_neexdzii_1968.csv")
table(sel$scale)
# 1:12000 1:31680
#      32       8
```

## What comes next

Order the photos from the
[BC Air Photo Warehouse](https://www2.gov.bc.ca/gov/content/data/geographic-data-services/air-photos),
scan or georeference them, then compare against modern satellite imagery
from [drift](https://github.com/NewGraphEnvironment/drift) to quantify
land cover change within the floodplain. The combination answers: what
changed, where, and when — the foundation for restoration prioritization.

## Ecosystem

| Package | Role in this workflow |
|---------|---------------------|
| [flooded](https://github.com/NewGraphEnvironment/flooded) | Delineated the floodplain AOI |
| [fly](https://github.com/NewGraphEnvironment/fly) | Computed footprints and ran coverage selection |
| **diggs** | Interactive exploration and export |
| [drift](https://github.com/NewGraphEnvironment/drift) | Next step — satellite land cover change analysis |
